---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Lilly McClendon"
date: "02-21-2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r load-Whickham-data}
data(Whickham)

?Whickham

#library(performance)
#performance::compare_performance()
```

### Exercise 1

I think these data came from an observational study because it is described as surveying participants between 1972 and 1974 and then following up 20 years later. Observational studies frequently involve researchers monitoring participants over time. Unlike in an experimental study, in this study, there was no manipulation of variables by the researcher or a control group. Observational studies are commonly used in public health or epidemiology research where the goal is to understand a potential association between an exposure and outcome. Further I think that this is an observational study because it would not be ethical to manipulate smoking knowing the risks to see the potential disease outcomes.  

### Exercise 2

```{r Whickham-view}
view(Whickham)
nrow(Whickham)
ncol(Whickham)
```

There are 1,314 observations in this data set. Each observation represents a subset of the survey sample and included UK women who either reported never smoked or currently smoked. 

### Exercise 3

There are 3 variables in the data set. Smoking (dichotomous, categorical) and age (continuous) can be considered explanatory variables and outcome can be considered a response variable (dichotomous, categorical). In this study, smoking is observed to see if there is an association with mortality. 


```{r smoking-visualizations}
library(ggplot2)
ggplot(Whickham, aes(x = smoker)) + 
  geom_bar() + 
  labs(title = "Smoking Status")
```

```{r mortality-visualizations}
library(ggplot2)
ggplot(Whickham, aes(x = outcome)) + 
  geom_bar() + 
  labs(title = "Mortality Outcomes")
```

```{r age-visualization}
library(ggplot2)
ggplot(Whickham, aes(x=age)) + 
  geom_histogram(binwidth = 2, color = "navy", fill = "grey") + 
  labs(title = "Age of Participants", x = "Age (years)") + 
  theme_minimal()
```

### Exercise 4

I would expect the relationship between smoking status and health outcome to be that more individuals who smoked have died. 

### Exercise 5

```{r smoking-outcome-visualization}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(prop_outcome = n / sum(n))

ggplot(Whickham, aes(y = smoker, fill = outcome)) + 
  geom_bar(position = "fill") + 
  labs(title = "Outcome by Smoking Status", 
       y = NULL, x = NULL)
```

The relationship between smoking status and health outcome did not meet my expectation that individuals who smoked were more likely to have died compared to individuals that did not smoke. The probability of being alive was highst for individuals who smoked, and individuals who didn't smoke had a higher probability of being deceased. 

### Exercise 6

```{r new-variable-age_cat}
library(dplyr)
Whickham <- Whickham %>% 
  mutate(
    age_cat = dplyr::case_when(
      age <= 44 ~ "18-44", 
      age > 44 & age <= 64 ~ "45-64",
      age > 64 ~ "65+"
  )
)
```

### Exercise 7

```{r re-created-visualization}
ggplot(Whickham, aes(y = smoker, fill = outcome)) + 
  geom_bar(position = "fill") + 
  facet_wrap(. ~ age_cat) +
  labs(title = "Outcome by Smoking Status and Age Category", 
       y = NULL, x = NULL)

Whickham %>%
  count(smoker, age_cat, outcome) %>%
  group_by() %>%
  mutate(prop_outcome = n / sum(n))
```

When looking at the recreated visualization showing the relationship between smoking status and health outcome faceted by age category, among all groups, it appears that across all age groups when individuals smoke, more of them are deceased. It appears to be particularly pronounced in the 45-64 age range which may be as that is when diseases from smoking may arise and lead to early mortality. It may be that when you examine the 65+ category there is not as dramatic a difference in mortality outcome between smokers and non-smokers as other health problems tend to arise in older adults potentially leading to their death.  