---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Lilly McClendon"
date: "03-02-2025"
output: github_document
---

### Load packages and data 
```{r load-packages, message = FALSE}
library(tidyverse) 

```

```{r load-data-instructional-staff}
staff <- read_csv("data/instructional-staff.csv")
view(staff)

# reshape the data 

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```
### Exercise 1

```{r line-plot-staff-long}
library(ggplot2)
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) + 
  labs(title = "Instructional Staff Employee Hires from 1975 - 2011", x = "Year", y = "Percentage of Hires") + 
  geom_line() + 
  guides(color = guide_legend(title = "Instructional Staff Type"))
```

### Exercise 2 

If the objective of this plot was to show that the proportion of part time faculty have gone up over time compared to other instructional staff types, I would use a stacked bar chart beccause it would allow for the proportion of eachs staff type to be compared easily across the years. A stacked bar chart would allow for both year and staff type to be compared with proportion together. Year would be the total bar, and the sub-bars would be the different staff types. 

### Exercise 3 

#### Improvement Ideas 

I would plan to improve the plot by making it a bar plot so that it would be easy to see the different subgroups of the data as the x axis is country which is a categorical variable and the goal is to compare different groups. Specifically, I would use a stacked bar chart as it would allow for an easier comparison of the fish captured and fish caught for each country and between countries. I will filter the data so that same to the example, only countries with a total harvest of over 100,000 will be included and only the countries labelled in the histogram will be included. I will change the y-axis to be logarithmic because the data has a very large range of values and it will make the data better able to be placed in a visual representation. The logarithmic data results in the data appearing more even so that the smaller harvest values (majority of the data) can still be seen . 

#### Changes 

* Mutated the dataset to add a column for the log value of capture and aquaculture.
* Changed plot to a stacked bar plot 


``` {r fisheries-improved-visualization}
library(dplyr)
library(ggplot2)

fisheries <- read_csv("data/fisheries.csv")

fisheries_filtered <- fisheries %>% 
  filter(total >= 100000) %>%
  filter(country == "China" | country == "Russia" | country == "Norway" | country == "Malaysia" | country == "Spain" | country == "Nigeria" | country == "France" | country == "Mauritania" | country == "Namibia" | country == "Ghana" | country == "Mozambique" | country == "Poland" | country == "Sierra Leone" | country == "Laos" | country == "Panama"| country == "Seychelles" | country == "Saudi Arabia") %>% 
  mutate(log_capture = log(capture + 1)) %>% 
  mutate(log_aquaculture = log(aquaculture + 1))
print(fisheries_filtered)

fisheries_filtered_long <- fisheries_filtered %>% 
  pivot_longer(cols = starts_with("log_capture") | starts_with("log_aquaculture"), 
               names_to = "fishing_method", 
               values_to = "value")

ggplot(fisheries_filtered_long, aes(x =reorder(country, -value), y= value, fill = fishing_method)) + 
  geom_bar(stat="identity") + 
  labs(title = "Fish Harvest by Country and Method ", x = "Country", y = " Fish Harvest (log(Tonnage))") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5)) + 
  scale_fill_discrete(name = "Harvesting Method", labels = c("Capture", "Aquaculture"))
```



## Stretch Practice 

### Load packages and data

```{r load-packages-stretch, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r load-Whickham-data}
data(Whickham)

?Whickham

#library(performance)
#performance::compare_performance()
```

### Stretch Practice 1

I think these data came from an observational study because it is described as surveying participants between 1972 and 1974 and then following up 20 years later. Observational studies frequently involve researchers monitoring participants over time. Unlike in an experimental study, in this study, there was no manipulation of variables by the researcher or a control group. Observational studies are commonly used in public health or epidemiology research where the goal is to understand a potential association between an exposure and outcome. Further I think that this is an observational study because it would not be ethical to manipulate smoking knowing the risks to see the potential disease outcomes.  

### Stretch Practice 2

```{r Whickham-view}
view(Whickham)
nrow(Whickham)
ncol(Whickham)
```

There are 1,314 observations in this data set. Each observation represents a subset of the survey sample and included UK women who either reported never smoked or currently smoked. 

### Stretch Practice 3

There are 3 variables in the data set. Smoking (dichotomous, categorical) and age (continuous) can be considered explanatory variables and outcome can be considered a response variable (dichotomous, categorical). In this study, smoking is observed to see if there is an association with mortality. 


```{r smoking-visualizations}
library(ggplot2)
ggplot(Whickham, aes(x = smoker)) + 
  geom_bar() + 
  labs(title = "Smoking Status")
```

```{r mortality-visualizations}
library(ggplot2)
ggplot(Whickham, aes(x = outcome)) + 
  geom_bar() + 
  labs(title = "Mortality Outcomes")
```

```{r age-visualization}
library(ggplot2)
ggplot(Whickham, aes(x=age)) + 
  geom_histogram(binwidth = 2, color = "navy", fill = "grey") + 
  labs(title = "Age of Participants", x = "Age (years)") + 
  theme_minimal()
```

### Stretch Practice 4

I would expect the relationship between smoking status and health outcome to be that more individuals who smoked have died. 

### Stretch Practice 5

```{r smoking-outcome-visualization}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(prop_outcome = n / sum(n))

ggplot(Whickham, aes(y = smoker, fill = outcome)) + 
  geom_bar(position = "fill") + 
  labs(title = "Outcome by Smoking Status", 
       y = NULL, x = NULL)
```

The relationship between smoking status and health outcome did not meet my expectation that individuals who smoked were more likely to have died compared to individuals that did not smoke. The probability of being alive was highst for individuals who smoked, and individuals who didn't smoke had a higher probability of being deceased. 

### Stretch Practice 6

```{r new-variable-age_cat}
library(dplyr)
Whickham <- Whickham %>% 
  mutate(
    age_cat = dplyr::case_when(
      age <= 44 ~ "18-44", 
      age > 44 & age <= 64 ~ "45-64",
      age > 64 ~ "65+"
  )
)
```

### Stretch Practice 7

```{r re-created-visualization}
ggplot(Whickham, aes(y = smoker, fill = outcome)) + 
  geom_bar(position = "fill") + 
  facet_wrap(. ~ age_cat) +
  labs(title = "Outcome by Smoking Status and Age Category", 
       y = NULL, x = NULL)

Whickham %>%
  count(smoker, age_cat, outcome) %>%
  group_by() %>%
  mutate(prop_outcome = n / sum(n))
```

When looking at the recreated visualization showing the relationship between smoking status and health outcome faceted by age category, among all groups, it appears that across all age groups when individuals smoke, more of them are deceased. It appears to be particularly pronounced in the 45-64 age range which may be as that is when diseases from smoking may arise and lead to early mortality. It may be that when you examine the 65+ category there is not as dramatic a difference in mortality outcome between smokers and non-smokers as other health problems tend to arise in older adults potentially leading to their death.  